---
title: "000_notes"
author: "Michael-Philipp Stiebing"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2022-12-05

* Download source data from **https://divvy-tripdata.s3.amazonaws.com/index.html**
* License for using the data from **https://ride.divvybikes.com/data-license-agreement**
* Create project root folder - **/20221205-capstone_datascience/** - as **../** in this document
* Move raw, zipped data to **../099_original_raw_data/001_raw_compressed/**
* Unzip data into **../099_original_raw_data/002_raw_csv/**

* Download copy of the assignment to \
**../098_documentation/20221205-DAC8_Case_Study_1.pdf**
* Download copy of the license to \
**../098_documentation/20221205-Data_License_Agreement_Divvy_Bikes.pdf**
* Download copy of the R Script to \
**../098_documentation/20221205-Copy_of_Divvy_Exercise_R_Script.txt**

* Create **../001_deliverables/**

* Copy data to **../002_data/001_csv/** as a working copy
* Rename file **202209-divvy-publictripdata.csv** in **../002_data/001_csv/** to **202209-divvy-tripdata.csv** in order to match filename schema

* Download quarterly formatted data from **https://divvy-tripdata.s3.amazonaws.com/index.html** to compare with monthly formatted data
* Copy dataset **Divvy_Trips_2020_Q1.csv** to **../099_original_raw_data/002_raw_csv/** and working copy to **../002_data/001_csv/**

* Comparing data in **Divvy_Trips_2020_Q1.csv** to **202210-divvy-tripdata.csv** as to similarity.
  + Using OpenOffice Calc V 4.1.7 for comparison
  + Column Names match
  + The rideable_type column contains more information in the **202210-divvy-tripdata.csv** as opposed to the **Divvy_Trips_2020_Q1.csv**
    + In **Divvy_Trips_2020_Q1.csv** only one type: dockable_bike
    + In **202210-divvy-tripdata.csv** either classic_bike or electric_bike
  + The columns start_station_id and end_station_id have different values
    + In **Divvy_Trips_2020_Q1.csv** the values are numerical
    + In **202210-divvy-tripdata.csv** some values are numerical, others are a mix of uppercase letters and numbers a few also include a dash '-'
  + The columns start_lat, start_lng, end_lat, end_lng are formatted differently
    + In **Divvy_Trips_2020_Q1.csv** the values are truncated to 4 decimal places
    + In **202210-divvy-tripdata.csv** the values contain up to 10 decimal places
  + Some records in **202210-divvy-tripdata.csv** do not contain start or end station data, leading me to believe it includes data from bikes that were rented not from a station, but standing around in the city.
  
* Checking **202110-divvy-tripdata.csv**, the oldest of the monthly datasets, to confirm that the differences persist
* I will be using the monthly datasets for the analysis, because:
  + The data is more current
  + It includes additional data as to the type of bike used
  + The data that is missing (station names) or not formatted nicely (gps coordinates) doesn't seem to impact the analysis

## 2022-12-06

* Creating R Studio Project in **../003_Rstudio_project/20221206-RStudio_Project01/**
* Create R Studio Script **001_Capstone01.R**
* Work through **20221205-Copy_of_Divvy_Exercise_R_Script** and adapt some commands to the new dataset

```{r load libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readr)
```

```{r create dataframe,message=FALSE, warning=FALSE}
setwd("/home/mikiR/remote_transfer/")
X202110_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202110-divvy-tripdata.csv")
X202111_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202111-divvy-tripdata.csv")
X202112_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202112-divvy-tripdata.csv")
X202201_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202201-divvy-tripdata.csv")
X202202_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202202-divvy-tripdata.csv")
X202203_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202203-divvy-tripdata.csv")
X202204_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202204-divvy-tripdata.csv")
X202205_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202205-divvy-tripdata.csv")
X202206_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202206-divvy-tripdata.csv")
X202207_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202207-divvy-tripdata.csv")
X202208_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202208-divvy-tripdata.csv")
X202209_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202209-divvy-tripdata.csv")
X202210_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202210-divvy-tripdata.csv")

all_trips <- bind_rows(X202110_divvy_tripdata, X202111_divvy_tripdata, X202112_divvy_tripdata, 
                       X202201_divvy_tripdata, X202202_divvy_tripdata, X202203_divvy_tripdata, 
                       X202204_divvy_tripdata, X202205_divvy_tripdata, X202206_divvy_tripdata, 
                       X202207_divvy_tripdata, X202208_divvy_tripdata, X202209_divvy_tripdata, 
                       X202210_divvy_tripdata)
```

```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
```

## 2023-01-09

* Continue preparing the dataset

* Remove bad data


```{r}
all_trips_v2 <- all_trips[!(all_trips$ride_length<0),]
```

* move bad data into a dataframe to doublecheck

```{r}
all_trips_errors <- all_trips[(all_trips$ride_length<0),]
```

* Conduct descriptive analysis as per script

```{r}

# Descriptive analysis on ride_length (all figures in seconds)
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride


# You can condense the four lines above to one line using summary() on the specific attribute
summary(all_trips_v2$ride_length)


# Compare members and casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)


# See the average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)


# Notice that the days of the week are out of order. Let's fix that.
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))


# Now, let's run the average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)


# analyze ridership data by type and weekday
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()                                                        #calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>%                 # calculates the average duration
  arrange(member_casual, weekday)                                                                # sorts


# Let's visualize the number of rides by rider type
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")


# Let's create a visualization for average duration
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")



```
## 2023-01-10

*   Trying to calculate the distance between gps coordinates strart / end

```{r}

library(geosphere) 
```

```{r}

all_trips_v3 <- all_trips_v2 %>%  
  mutate(geodist = distHaversine(cbind(all_trips_v2$start_lng,all_trips_v2$start_lat),cbind(all_trips_v2$end_lng,all_trips_v2$end_lat)))
```


*   To fix performance issues, I had to setup a RStudio server 

          CentOS 9 Stream 
          RStudio 2022.12.0+353 "Elsbeth Geranium" Release
          R version 4.2.2



```{r eval=FALSE, include=FALSE}

all_trips_v3 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(average_distance = mean(geodist)
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = geodist, fill = member_casual)) +
  geom_col(position = "dodge")

```

## 2023-01-17

*   filter all trips with distance = 0 into a dataframe called round_trips, the assumtion being that when the trip ends where it started

```{r}
round_trips = filter(all_trips_v3,geodist == 0)
```

```{r}

# Descriptive analysis on ride_length (all figures in seconds)
mean(round_trips$ride_length) #straight average (total ride length / rides)
median(round_trips$ride_length) #midpoint number in the ascending array of ride lengths
max(round_trips$ride_length) #longest ride
min(round_trips$ride_length) #shortest ride


# You can condense the four lines above to one line using summary() on the specific attribute
summary(round_trips$ride_length)


# Compare members and casual users
aggregate(round_trips$ride_length ~ round_trips$member_casual, FUN = mean)
aggregate(round_trips$ride_length ~ round_trips$member_casual, FUN = median)
aggregate(round_trips$ride_length ~ round_trips$member_casual, FUN = max)
aggregate(round_trips$ride_length ~ round_trips$member_casual, FUN = min)


# See the average ride time by each day for members vs casual users
aggregate(round_trips$ride_length ~ round_trips$member_casual + round_trips$day_of_week, FUN = mean)


# Notice that the days of the week are out of order. Let's fix that.
round_trips$day_of_week <- ordered(round_trips$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))


# Now, let's run the average ride time by each day for members vs casual users
aggregate(round_trips$ride_length ~ round_trips$member_casual + round_trips$day_of_week, FUN = mean)


# analyze ridership data by type and weekday
round_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()                                                        #calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>%                 # calculates the average duration
  arrange(member_casual, weekday)                                                                # sorts


# Let's visualize the number of rides by rider type
round_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")


# Let's create a visualization for average duration
round_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```
```{r}
aggregate(round_trips$ride_length ~ round_trips$member_casual, FUN = mean)

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v3$ride_length ~ all_trips_v3$member_casual, FUN = mean)
aggregate(round_trips$ride_length ~ round_trips$member_casual, FUN = mean)
aggregate(round_trips$ride_length ~ round_trips$member_casual + round_trips$day_of_week, FUN = mean)

```
## 2023-02-22
```{r}
# Let's create a visualization for average duration
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = rideable_type)) +
  geom_col()
```


```{r}
round_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  filter(member_casual == 'casual') %>%
  group_by(weekday,rideable_type) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = rideable_type)) +
  geom_col()
```

## 2023-03-06

* While working with the data, I noticed some very long rental periods ( > 86400 seconds, aka one day) I think this would be maintainance work, as this would be unlikely to be a rental from a customer
* Also, I filtered out all rentals under 1 minute, since these do not represent useful customer behavior as well
* Similarly, all entries that have a geodist value of NA - because they don't have a return station value, are similarly out for maintainance, or stolen, or in case of very short rental periods of a few seconds, maybe a booking error or system test

* I filtered these sets into dataframes into short_ride long_ride and nadist_ride in order to analyze a little bit

```{r}
short_ride <- (filter(all_trips_v3,ride_length < 60 & !is.na(geodist)))
long_ride <- (filter(all_trips_v3,ride_length > 86400 & !is.na(geodist)))
nadist_ride <- (filter(all_trips_v3,is.na(geodist)))

summary(filter(short_ride)$ride_length)
summary(filter(long_ride)$ride_length)
summary(filter(nadist_ride)$ride_length)

all_trips_v4 <- (filter(all_trips_v3,ride_length < 86400 & ride_length > 60 & ride_length != 0 & !is.na(geodist)))
```


## 2023-03-07

* Trying to find out more about the different behaviours between members and casual users
* First I tried to compare rides per group that cross a certain Threshhold, I chose the 45 min (2700 seconds) and the 3 hour threshhold (10800 seconds)
* This seems to confirm: Casual users tend to take a lot longer rides compared to members
* Then I tried plotting instances of casual rides vs members rides across the 12 month period.
  + Total rides are highest in summer, lowest in winter
  + More members use the service than casual users, except on the weekends, especially in summer
  + This trend is even more pronounced for "round trips" where geodist = 0 This could be tourists



```{r}
nrow(all_trips_v4[all_trips_v4$member_casual == "casual",])
nrow(all_trips_v4[all_trips_v4$member_casual == "member",])
#nrow(all_trips_v4[all_trips_v4$member_casual == "casual" & all_trips_v4$ride_length > 2700,])
nrow(all_trips_v4[all_trips_v4$member_casual == "casual" & all_trips_v4$ride_length > 2700,]) / nrow(all_trips_v4[all_trips_v4$member_casual == "casual",])
nrow(all_trips_v4[all_trips_v4$member_casual == "member" & all_trips_v4$ride_length > 2700,]) / nrow(all_trips_v4[all_trips_v4$member_casual == "member",])
nrow(all_trips_v4[all_trips_v4$member_casual == "casual" & all_trips_v4$ride_length > 10800,]) / nrow(all_trips_v4[all_trips_v4$member_casual == "casual",])

all_trips_v4 %>% 
  group_by(month, member_casual) %>%
  summarize(cascount = sum(member_casual == "casual"), memcount = sum(member_casual == "member"))


ggplot(all_trips_v4, aes(date, color = member_casual)) + 
  geom_freqpoly(binwidth=7)


all_trips_v4 %>% 
  filter(day_of_week == "Saturday" | day_of_week == "Sunday") %>%
ggplot(aes(date, color = member_casual)) + 
  geom_freqpoly(binwidth=7)

all_trips_v4 %>% 
  filter(day_of_week != "Saturday" | day_of_week != "Sunday") %>%
ggplot(aes(date, color = member_casual)) + 
  geom_freqpoly(binwidth=7)

all_trips_v4 %>% 
  filter(geodist == 0) %>%
  filter(day_of_week == "Saturday" | day_of_week == "Sunday") %>%
ggplot(aes(date, color = member_casual)) + 
  geom_freqpoly(binwidth=7)


all_trips_v4 %>% 
  filter(geodist == 0) %>%
ggplot(aes(date, color = member_casual)) + 
  geom_freqpoly(binwidth=7)
# 
# all_trips_v4 %>% 
#   filter(day_of_week == "Monday") %>%
# ggplot(aes(date, color = member_casual)) + 
#   geom_freqpoly(binwidth=7)
# 
# all_trips_v4 %>% 
#   filter(day_of_week == "Tuesday") %>%
# ggplot(aes(date, color = member_casual)) + 
#   geom_freqpoly(binwidth=7)
# 
# all_trips_v4 %>% 
#   filter(day_of_week == "Wednesday") %>%
# ggplot(aes(date, color = member_casual)) + 
#   geom_freqpoly(binwidth=7)
# 
# all_trips_v4 %>% 
#   filter(day_of_week == "Thursday") %>%
# ggplot(aes(date, color = member_casual)) + 
#   geom_freqpoly(binwidth=7)
# 
# all_trips_v4 %>% 
#   filter(day_of_week == "Friday") %>%
# ggplot(aes(date, color = member_casual)) + 
#   geom_freqpoly(binwidth=7)
# 

# 2700s > 45 min 10800s > 3 hours

  
```
## 2023-03-09

+ Trying to poke around with gps coordinates, gave up because its too time consuming for this capstone project

```{r}
library(plotKML)
library(ggplot2)
install.packages("ggmap")
library(ggmap)
library(RColorBrewer)
```
```{r}
all_trips_v5 %>% 
  filter(geodist == 0) %>%
 ggplot(aes(x = start_lng, y = start_lat)) +
  coord_quickmap() +
  geom_point()

all_trips_v5 %>% 
 ggplot(aes(x = start_lng, y = start_lat)) +
  coord_quickmap() +
  geom_point()
```
## 2023-03-10

+ Trying to find out if speed is plausible and filter out outliers

```{r}

nrow(all_trips_v4[all_trips_v4$geodist > 30000,])

summary(all_trips_v4$geodist)

View(all_trips_v4[all_trips_v4$geodist > 30000,])

temp <- all_trips_v4 %>% 
  mutate(speed = geodist / ride_length) %>%
  filter(speed < 30)



all_trips_v4 %>% 
  mutate(speed = geodist / ride_length) %>%
  filter(speed > 15)

all_trips_v5 <- all_trips_v4 %>% 
  mutate(speed = geodist / ride_length) %>%
  filter(speed < 10)


sd(temp$speed)

sd(all_trips_v5$speed)

summary(all_trips_v5$speed)

hist(all_trips_v5$speed,breaks=20)
hist(temp$speed,breaks=20)
```


## 2023-03-15

+ Tried to find out at which ride length there is the biggest difference between members and casual users, its around 11 minutes


```{r}
variation <- data.frame(threshhold=integer(),
                        casual_percentage=double(),
                        member_percentage=double()
                        )
for ( i in seq(60,12000,60)) {
  new <- c(i,
            nrow(all_trips_v5[all_trips_v5$member_casual == "casual" & all_trips_v5$ride_length > i,]) / nrow(all_trips_v5[all_trips_v5$member_casual == "casual",]),
            nrow(all_trips_v5[all_trips_v5$member_casual == "member" & all_trips_v5$ride_length > i,]) / nrow(all_trips_v5[all_trips_v5$member_casual == "member",]))
  variation[nrow(variation) + 1,] <- new
}

variation_smaller <- data.frame(threshhold=integer(),
                        casual_percentage=double(),
                        member_percentage=double()
                        )
for ( i in seq(60,12000,60)) {
  new <- c(i,
            nrow(all_trips_v5[all_trips_v5$member_casual == "casual" & all_trips_v5$ride_length < i,]) / nrow(all_trips_v5[all_trips_v5$member_casual == "casual",]),
            nrow(all_trips_v5[all_trips_v5$member_casual == "member" & all_trips_v5$ride_length < i,]) / nrow(all_trips_v5[all_trips_v5$member_casual == "member",]))
  variation_smaller[nrow(variation_smaller) + 1,] <- new
}

ggplot(variation, aes(x=threshhold)) + 
 # geom_line(aes(y = casual_percentage), color = "darkred") + 
#  geom_line(aes(y = member_percentage), color="steelblue", linetype="twodash") +
  geom_line(aes(y = casual_percentage-member_percentage), color="black", linetype="twodash")+ xlim(500,1000)

ggplot(variation_smaller, aes(x=threshhold)) + 
  geom_line(aes(y = casual_percentage), color = "darkred") + 
  geom_line(aes(y = member_percentage), color="steelblue", linetype="twodash") +
  geom_line(aes(y = casual_percentage-member_percentage), color="black", linetype="twodash")
#+
# xlim(500,1000)

nrow(all_trips_v5[all_trips_v5$member_casual == "casual" & all_trips_v5$ride_length < 120,]) / nrow(all_trips_v5[all_trips_v5$member_casual == "casual",])
```

+ misc backup of old visualizations

```{r}
all_trips_v5 %>% 
  filter(day_of_week != "Saturday" | day_of_week != "Sunday" ) %>%
  ggplot( aes(x = date, fill = member_casual)) +
  geom_density(alpha = .3) +
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = NULL, x = NULL, fill = NULL)


all_trips_v5 %>% 
  filter(day_of_week == "Saturday" | day_of_week == "Sunday" ) %>%
  ggplot( aes(x = date, fill = member_casual)) +
  geom_density(alpha = .3) +
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = NULL, x = NULL, fill = NULL)


all_trips_v5 %>% 
  filter(day_of_week == "Saturday" | day_of_week == "Sunday" ) %>%
  group_by(member_casual, hours) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, hours)  %>% 
  ggplot( aes(x = hours, y = number_of_rides / 1000, fill = member_casual)) +
  geom_col(position = "dodge") + 
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = "number of rides in thousands", x = NULL, fill = NULL)

all_trips_v5 %>% 
  filter(day_of_week != "Saturday" | day_of_week != "Sunday" ) %>%
  group_by(member_casual, hours) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, hours)  %>% 
  ggplot( aes(x = hours, y = number_of_rides / 1000, fill = member_casual)) +
  geom_col(position = "dodge") + 
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = "number of rides in thousands", x = NULL, fill = NULL)


```

## 2023-03-20

* Tried visualizing usage by hour of the day as a radar chart, ultimately, I think the barchart is clearer and easier to understand


```{r}
install.packages("fmsb")
library(fmsb)


members_numrides <- all_trips_v5 %>% 
  filter(member_casual=="member") %>%
  group_by(group = as.numeric(hours) ) %>%
  summarise(number_of_rides = n()) 
casuals_numrides <- all_trips_v5 %>% 
  filter(member_casual=="casual") %>%
  group_by(group = as.numeric(hours) ) %>%
  summarise(number_of_rides = n()) 

numrides_by_hour <- data.frame("00"=integer(),
                        "01"=integer(),
                        "02"=integer(),
                        "03"=integer(),
                        "04"=integer(),
                        "05"=integer(),
                        "06"=integer(),
                        "07"=integer(),
                        "08"=integer(),
                        "09"=integer(),
                        "10"=integer(),
                        "11"=integer(),
                        "12"=integer(),
                        "13"=integer(),
                        "14"=integer(),
                        "15"=integer(),
                        "16"=integer(),
                        "17"=integer(),
                        "18"=integer(),
                        "19"=integer(),
                        "20"=integer(),
                        "21"=integer(),
                        "22"=integer(),
                        "23"=integer()
                        )

if (max(t(members_numrides)[2,]) > max(t(casuals_numrides)[2,])) {
  numrides_by_hour[nrow(numrides_by_hour) + 1,] <- rep(max(t(members_numrides)[2,]),ncol(numrides_by_hour))  
} else {
  numrides_by_hour[nrow(numrides_by_hour) + 1,] <- rep(max(t(casuals_numrides)[2,]),ncol(numrides_by_hour))
}

#if (min(t(members_numrides)[2,]) < min(t(casuals_numrides)[2,])) {
#  numrides_by_hour[nrow(numrides_by_hour) + 1,] <- rep(min(t(members_numrides)[2,]),ncol(numrides_by_hour))  
#} else {
#  numrides_by_hour[nrow(numrides_by_hour) + 1,] <- rep(min(t(casuals_numrides)[2,]),ncol(numrides_by_hour))
#}
numrides_by_hour[nrow(numrides_by_hour) + 1,] <- rep(0,ncol(numrides_by_hour))

numrides_by_hour[nrow(numrides_by_hour) + 1,] <- t(members_numrides)[2,]
numrides_by_hour[nrow(numrides_by_hour) + 1,] <- t(casuals_numrides)[2,]


radarchart(numrides_by_hour)

variation[nrow(variation)+1,] <- (rep(max(tempos),24))

variation[nrow(variation)+1,] <- (rep(min(tempos),24))

variation[nrow(variation)+1,] <- (tempos[2,])

tempost=as.data.frame(tempos[2,])

radarchart(variation)  
  
```

* Tried visualizing the correlation of temperature and usage

```{r}

dates_vline <- as.Date(c("2022-01-17","2022-02-12","2022-02-21","2022-03-07","2022-05-30","2022-06-19","2022-07-04","2022-09-05","2022-10-10"))      
dates_vline <- which(all_trips_v5$date %in% dates_vline)


all_trips_v5 %>% 
  ggplot() +
#  geom_line(data = climate_chicago_202110_202210, aes(x = date, y = avgtemp / 20000), color = "red", linetype = "dashed",alpha=.5) +
  geom_smooth(data = climate_chicago_202110_202210, aes(x = date, y = avgtemp / 20000),method=loess,color="#D55E00") +
  geom_density( aes(x = date,fill = member_casual),alpha = .3) +
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = NULL, x = NULL, fill = NULL)
 # geom_vline(xintercept = as.numeric(all_trips_v5$date[dates_vline]), col = "red", lwd = 0.4)
  
  
  ggplot() +
  geom_line( aes(x = climate_chicago_202110_202210$date, y = climate_chicago_202110_202210$avgtemp))

  
  

#  geom_line(data = climate_chicago_202110_202210,  aes(x = date, y = avgtemp / 20000)) +

```