---
title: "Processing data sources for analysis"
author: "Michael-Philipp Stiebing"
date: "2023-03-10"
output: 
  pdf_document: 
    latex_engine: lualatex
    keep_tex: yes
mainfont: Bahnschrift:style=Bold
monofont: Bahnschrift:style=Condensed
---


```{r setup, include=FALSE }
knitr::opts_chunk$set(echo = TRUE)
```
\fontsize{10}{12}
\selectfont
* Start by importing libraries

\fontsize{9}{11}
\selectfont
```{r load libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readr)
```
\fontsize{10}{12}
\selectfont
* Read the csv files of the Divvy data into separate dataframes
* Combine all dataframes into a single dataframe

\fontsize{9}{11}
\selectfont
```{r create dataframe,message=FALSE, warning=FALSE}
setwd("/home/mikiR/remote_transfer/")
X202110_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202110-divvy-tripdata.csv")
X202111_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202111-divvy-tripdata.csv")
X202112_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202112-divvy-tripdata.csv")
X202201_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202201-divvy-tripdata.csv")
X202202_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202202-divvy-tripdata.csv")
X202203_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202203-divvy-tripdata.csv")
X202204_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202204-divvy-tripdata.csv")
X202205_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202205-divvy-tripdata.csv")
X202206_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202206-divvy-tripdata.csv")
X202207_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202207-divvy-tripdata.csv")
X202208_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202208-divvy-tripdata.csv")
X202209_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202209-divvy-tripdata.csv")
X202210_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202210-divvy-tripdata.csv")

all_trips <- bind_rows(
                  X202110_divvy_tripdata, X202111_divvy_tripdata, X202112_divvy_tripdata, X202201_divvy_tripdata,
                  X202202_divvy_tripdata, X202203_divvy_tripdata, X202204_divvy_tripdata, X202205_divvy_tripdata,
                  X202206_divvy_tripdata, X202207_divvy_tripdata, X202208_divvy_tripdata, X202209_divvy_tripdata, 
                  X202210_divvy_tripdata)

```
\fontsize{10}{12}
\selectfont

* Create new columns from date field, split into
    + month
    + day
    + year
    + day of week
    + time of day
    + hour

\fontsize{9}{11}
\selectfont
```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
all_trips$timeofday <- format(as.POSIXct(all_trips$started_at), format = "%H:%M:%S")
all_trips$hours <- format(as.POSIXct(all_trips$started_at), format = "%H")
```
\fontsize{10}{12}
\selectfont

* Create column ridelength calculated from the difference of ended_at and started_at
* Convert column value into numeric

\fontsize{9}{11}
\selectfont
```{r}

all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
```
\fontsize{10}{12}
\selectfont

* Remove bad data

\fontsize{9}{11}
\selectfont
```{r}
all_trips_v2 <- all_trips[!(all_trips$ride_length<0),]
```
\fontsize{10}{12}
\selectfont

* Move bad data into a dataframe to doublecheck

\fontsize{9}{11}
\selectfont
```{r}
all_trips_errors <- all_trips[(all_trips$ride_length<0),]
```

\fontsize{10}{12}
\selectfont

*   Read the climate data into a dataframe

\fontsize{9}{11}
\selectfont
```{r,message=FALSE, warning=FALSE}
setwd("/home/mikiR/remote_transfer/")
climate_chicago_202110_202210 <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/climate_chicago_202110-202210.csv")

```


\fontsize{10}{12}
\selectfont

*   Convert the date field in the climate data frame to date format

\fontsize{9}{11}
\selectfont
```{r}

climate_chicago_202110_202210$date <- as.Date(climate_chicago_202110_202210$date)
```

\fontsize{10}{12}
\selectfont

*   Trying to calculate the distance between gps coordinates start_lng & start_lat / end_lng & end_lat
*   Import geosphere library

\fontsize{9}{11}
\selectfont
```{r}

library(geosphere) 
```
\fontsize{10}{12}
\selectfont

*   Calculate the distance between the start and end coordinates using the Haversine formula
*   Create new column geodist with the calculated distance

\fontsize{9}{11}
\selectfont
```{r}

all_trips_v3 <- all_trips_v2 %>% 
              mutate(geodist = distHaversine(
                    cbind(all_trips_v2$start_lng,all_trips_v2$start_lat), 
                    cbind(all_trips_v2$end_lng,all_trips_v2$end_lat)))
```
\fontsize{10}{12}
\selectfont

*   Filter all trips with distance = 0 into a dataframe called round_trips, the assumtion being that when the trip ends where it started

\fontsize{9}{11}
\selectfont
```{r}
round_trips = filter(all_trips_v3,geodist == 0)
```
\fontsize{10}{12}
\selectfont

* While working with the data, I noticed some very long rental periods ( > 86400 seconds, aka one day) I think this would be maintenance work, as this would be unlikely to be a rental from a customer
* Also, I filtered out all rentals under 1 minute, since these do not represent useful customer behavior as well
* Similarly, all entries that have a geodist value of NA - because they don't have a return station value, are similarly out for maintenance, or stolen, or in case of very short rental periods of a few seconds, maybe a booking error or system test

* I filtered these sets into dataframes into short_ride long_ride and nadist_ride in order to analyze a little bit

\fontsize{9}{11}
\selectfont
```{r}
short_ride <- (filter(all_trips_v3,ride_length < 60 & !is.na(geodist)))
long_ride <- (filter(all_trips_v3,ride_length > 86400 & !is.na(geodist)))
nadist_ride <- (filter(all_trips_v3,is.na(geodist)))

summary(filter(long_ride)$ride_length)
summary(filter(nadist_ride)$ride_length)
```
\fontsize{10}{12}
\selectfont

* Create a new dataframe all_trips_v4 containing only trips under 86400 s & geodist not NA

\fontsize{9}{11}
\selectfont
```{r}
all_trips_v4 <- (filter(all_trips_v3,ride_length < 86400 & ride_length > 60 & ride_length != 0 & !is.na(geodist)))
summary(filter(all_trips_v4)$ride_length)

```
\fontsize{10}{12}
\selectfont

* Another oddity I found in the data was rides with very high speeds, when dividing geodist in meters by ridelength in seconds
* I created a new column speed with the calculated value geodist / ride_length, and filter out all rides with speed above 10 m/s (36 km/h or 22.4 mph) - pedal assisted bikes are capped at 20 mph
* If the value is very low, this can mean a leisurely ride, or a ride that ends close to where it began, whereas a high value can only mean transport by other means, and since rented bikes are not allowed on CTA (https://www.transitchicago.com/bikeandride/) this can only mean the bike was not being used by a customer
* looking at the distribution of values, the bulk of non round trips seems to be around the average biking speed, the mean of the speed column being at 2.9 m/s - 6.5 mph



\fontsize{9}{11}
\selectfont
```{r}
all_trips_v5 <- all_trips_v4 %>% 
  mutate(speed = geodist / ride_length) %>%
  filter(speed < 10)
summary(all_trips_v5$speed)
hist(all_trips_v5$speed)

```
\fontsize{10}{12}
\selectfont

* I decided to create a new dataframe, all_trips_v5, filtering out the entries with impausibly high speed