---
title: "Analysis of Cyclistic User Data"
author: "Michael-Philipp Stiebing"
date: "2023-03-21"
output: 
  pdf_document: 
    latex_engine: lualatex
    keep_tex: yes
mainfont: Bahnschrift:style=Bold
monofont: Bahnschrift:style=Condensed
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readr)
library(geosphere)

setwd("/home/mikiR/remote_transfer/")
X202110_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202110-divvy-tripdata.csv")
X202111_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202111-divvy-tripdata.csv")
X202112_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202112-divvy-tripdata.csv")
X202201_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202201-divvy-tripdata.csv")
X202202_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202202-divvy-tripdata.csv")
X202203_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202203-divvy-tripdata.csv")
X202204_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202204-divvy-tripdata.csv")
X202205_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202205-divvy-tripdata.csv")
X202206_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202206-divvy-tripdata.csv")
X202207_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202207-divvy-tripdata.csv")
X202208_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202208-divvy-tripdata.csv")
X202209_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202209-divvy-tripdata.csv")
X202210_divvy_tripdata <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/202210-divvy-tripdata.csv")

all_trips <- bind_rows(
                  X202110_divvy_tripdata, X202111_divvy_tripdata, X202112_divvy_tripdata, X202201_divvy_tripdata,
                  X202202_divvy_tripdata, X202203_divvy_tripdata, X202204_divvy_tripdata, X202205_divvy_tripdata,
                  X202206_divvy_tripdata, X202207_divvy_tripdata, X202208_divvy_tripdata, X202209_divvy_tripdata, 
                  X202210_divvy_tripdata)

climate_chicago_202110_202210 <- 
  read_csv("./20221205-capstone_datascience-01/002_data/001_csv/climate_chicago_202110-202210.csv")

```

\fontsize{9}{11}
\selectfont
```{r include=FALSE}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
all_trips$timeofday <- format(as.POSIXct(all_trips$started_at), format = "%H:%M:%S")
all_trips$hours <- format(as.POSIXct(all_trips$started_at), format = "%H")

all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))

all_trips_v2 <- all_trips[!(all_trips$ride_length<0),]
all_trips_errors <- all_trips[(all_trips$ride_length<0),]

all_trips_v3 <- all_trips_v2 %>% 
              mutate(geodist = distHaversine(
                    cbind(all_trips_v2$start_lng,all_trips_v2$start_lat), 
                    cbind(all_trips_v2$end_lng,all_trips_v2$end_lat)))

all_trips_v4 <- (filter(all_trips_v3,ride_length < 86400 & ride_length > 60 & ride_length != 0 & !is.na(geodist)))

all_trips_v5 <- all_trips_v4 %>% 
  mutate(speed = geodist / ride_length) %>%
  filter(speed < 10)

climate_chicago_202110_202210$date <- as.Date(climate_chicago_202110_202210$date)

```
\fontsize{10}{12}
\selectfont

# Starting with processed dataframe all_trips_v5

* Conduct descriptive analysis as per script

\fontsize{9}{11}
\selectfont
```{r}

# Set up two color palette that are compatible with all kinds of color vision, 
# from http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/

cbPaletteMin <- c("#E69F00", "#0072B2")

# 2700s = 45 min 10800s = 3 hours

table(all_trips_v5$member_casual)

summary(all_trips_v5$ride_length)

summary(all_trips_v5$geodist)

# Compare members and casual users
aggregate(all_trips_v5$ride_length ~ all_trips_v5$member_casual, FUN = mean)
aggregate(all_trips_v5$ride_length ~ all_trips_v5$member_casual, FUN = median)

# casual users seem to take longer rides


nrow(all_trips_v5[all_trips_v5$member_casual == "casual" & all_trips_v5$ride_length > 2700,]) /
  nrow(all_trips_v5[all_trips_v5$member_casual == "casual",])
nrow(all_trips_v5[all_trips_v5$member_casual == "member" & all_trips_v5$ride_length > 2700,]) / 
  nrow(all_trips_v5[all_trips_v5$member_casual == "member",])

# 10 percent of casual users take rides that are longer than 45 minutes, whereas only 1.7 percent of members do

nrow(filter(all_trips_v5,geodist <= 10)) / nrow(all_trips_v5)

nrow(all_trips_v5[all_trips_v5$member_casual == "casual" & all_trips_v5$geodist <= 10,]) / 
  nrow(all_trips_v5[all_trips_v5$member_casual == "casual",])
nrow(all_trips_v5[all_trips_v5$member_casual == "member" & all_trips_v5$geodist <= 10,]) / 
  nrow(all_trips_v5[all_trips_v5$member_casual == "member",])

# 7.3 percent of casual users 'take round trips', whereas only 3.6 percent of members do

# Notice that the days of the week are out of order. Let's fix that.
all_trips_v5$day_of_week <- ordered(all_trips_v5$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday",
                                                                       "Thursday", "Friday", "Saturday"))

# Now, let's run the average ride time by each day for members vs casual users
aggregate(all_trips_v5$ride_length ~ all_trips_v5$member_casual + all_trips_v5$day_of_week, FUN = mean)


# analyze ridership data by type and weekday
all_trips_v5 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%    # creates weekday field using wday()
  group_by(member_casual, weekday) %>%                    # groups by usertype and weekday
  summarise(number_of_rides = n()                         # calculates the number of rides and average duration
            ,average_duration = mean(ride_length)) %>%    # calculates the average duration
  arrange(member_casual, weekday)                         # sorts


# Let's visualize the number of rides by rider type
all_trips_v5 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()/1000
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + 
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = "Number of rides in thousands", x = NULL, fill = NULL)

# ggsave("weeklyOverview.png",width=3840,height = 2160,unit="px")

# Let's visualize the number of rides by rider type for rides that begin and end within 
# 10 meters of each other 'round trips'
filter(all_trips_v5,geodist <= 10) %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") + 
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = "Number of rides", x = NULL, fill = NULL)

# ggsave("weeklyOverview-roundTrips.png",width=3840,height = 2160,unit="px")

# Let's create a visualization for average duration
all_trips_v5 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length / 60)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + 
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = "average trip duration in minutes", x = NULL, fill = NULL)

# ggsave("weeklyOverview-duration.png",width=3840,height = 2160,unit="px")

# Let's create a visualization for average duration  for rides that begin and end within 
# 10 meters of each other 'round trips'
filter(all_trips_v5,geodist <= 10) %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length / 60)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + 
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = "average trip duration in minutes", x = NULL, fill = NULL)

```
\fontsize{10}{12}
\selectfont

* The pattern emerging seems to be the following:
  + Members take shorter rides
  + Casual users take longer rides, especially on weekends
  + Casual users take more round trips than members


\fontsize{9}{11}
\selectfont
```{r}

# Let's create a visualization for number of rides by User type, over the whole timeframe. 
# Overlay average daily temperature in Fahrenheit. ( Has to be scaled to match the geom_density plot)
all_trips_v5 %>% 
  ggplot() +
  geom_smooth(data = climate_chicago_202110_202210, aes(x = date, y = as.double(avgtemp) / 20000),method=loess,color="#D55E00") +
  geom_density( aes(x = date,fill = member_casual),alpha = .3) +
  scale_y_continuous(
    breaks=c(seq(0.000,0.005,by=0.0005)), 
    labels=c(seq(0,100,10)),
    sec.axis = sec_axis(~ (((. * 20000) - 32) * 5/9), name = "Average daily temperature in °C",breaks=c(seq(-15,40,5)))
    ) +
  scale_x_date(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = "Average daily temperature in °F", x = NULL, fill = NULL) 

# ggsave("yearlyOverview.png",width=3840,height = 2160,unit="px")


```
\fontsize{10}{12}
\selectfont

* Looking at usage across the year, during the summer months, casual users overtake members
* Usage, especially by casual users, seems to correlate with average temperature

\fontsize{9}{11}
\selectfont

```{r}

# Let's create a visualization for number of rides by User type, only looking at time of day
all_trips_v5 %>% 
  group_by(member_casual, hours) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual, as.numeric(hours))  %>% 
  ggplot( aes(x = as.numeric(hours), y = number_of_rides / 1000, fill = member_casual)) +
  geom_col(position = "dodge") + 
  coord_flip()  +
  scale_x_reverse(breaks = (0:23)) +
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = "Number of rides in thousands", x = "Time of day", fill = NULL)

# ggsave("dailyOverview.png",width=3840,height = 2160,unit="px")

all_trips_v5 %>% 
  filter(day_of_week == "Saturday" | day_of_week == "Sunday" ) %>%
  group_by(member_casual, hours) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual, as.numeric(hours))  %>% 
  ggplot( aes(x = as.numeric(hours), y = number_of_rides / 1000, fill = member_casual)) +
  geom_col(position = "dodge") + 
  coord_flip()  +
  scale_x_reverse(breaks = (0:23)) +
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = "number of rides in thousands", x = NULL, fill = NULL)


all_trips_v5 %>% 
  filter(day_of_week != "Saturday" | day_of_week != "Sunday" ) %>%
  group_by(member_casual, hours) %>% 
  summarise(number_of_rides = n()) %>% 
  arrange(member_casual, as.numeric(hours))  %>% 
  ggplot( aes(x = as.numeric(hours), y = number_of_rides / 1000, fill = member_casual)) +
  geom_col(position = "dodge") + 
  coord_flip()  +
  scale_x_reverse(breaks = (0:23)) +
  scale_fill_manual(values=cbPaletteMin) +
  labs(y = "number of rides in thousands", x = NULL, fill = NULL)


```
\fontsize{10}{12}
\selectfont

* Looking at usage over hours of the day, members are clustered around morning and afternoon, preceding the start of a typical workday and following the end of it, which seems to support the theory that members mainly use the service to commute


